---
title: "Book Recommendation System"
output: html_document
date: "2024-03-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
For the initial phase of your project involving the construction of a book recommendation system, you are focusing on conducting an Exploratory Data Analysis (EDA). This fundamental step will involve a comprehensive examination and analysis of the dataset's features or variables. 

As part of our project, we are developing a book recommendation system. We have sourced our dataset from https://www.kaggle.com/datasets/arashnic/book-recommendation-dataset/code?datasetId=1004280&searchQuery=EDA

The first phase of our project is dedicated to performing an Exploratory Data Analysis (EDA). This process consists of  a thorough examination and analysis of the dataset's variables. Our EDA aims to explore the distribution of variables, investigate the interactions among them, identify and analyze outliers, and address missing values. We will conduct uni-variate, bi-variate, and multi-variate analyses to gain a deep understanding of the data, setting a solid foundation for developing an effective recommendation system.

So let us propose the agenda of our work pertaining to EDA:

1. Printing the first 5 rows of dataframe
2. Display the size of each dataframe
3. Inspection of null values present.
4. 

Let's load in any required packages here (make sure to install them first)
```{r loadingPackages}
require(dplyr)
require(readr)
require(caret)
require(pROC)

set.seed(1)

books = read_csv('/Users/tejk/Desktop/AA/task 3/datasets/Books.csv')
ratings  = read_csv('/Users/tejk/Desktop/AA/task 3/datasets/Ratings.csv')
users=read_csv('/Users/tejk/Desktop/AA/task 3/datasets/Users.csv')
```

Let us print the first five rows of each dataframe:

```{r}
#Printing the first five rows of books dataframe
head(books, n = 5)
```
```{r}
#Printing the first five rows of books dataframe
head(ratings, n = 5)
```

```{r}
#Printing the first five rows of books dataframe
head(users, n = 5)
```
2. Printing the size of each dataframe

```{r}


# Or using nrow() and ncol() for a more detailed approach
print(paste("Ratings: Rows =", nrow(ratings), "Columns =", ncol(ratings)))
print(paste("Users: Rows =", nrow(users), "Columns =", ncol(users)))
print(paste("Books: Rows =", nrow(books), "Columns =", ncol(books)))

```
3. Let us inspect the Ratings dataset. We shall visualize the following: Rating Distribution and Average Ratings per Book
But before that let us get the percentage of missing values for ratings dataset.

```{r}

total_rows <- nrow(ratings)
missing_values <- colSums(is.na(ratings))
percentage_missing <- (missing_values / total_rows) * 100

# Display the percentage of missing values for each column
print(percentage_missing)

```
There are no missing values for ratings data!

```{r}

total_rows <- nrow(users)
missing_values <- colSums(is.na(users))
percentage_missing <- (missing_values / total_rows) * 100

# Display the percentage of missing values for each column
print(percentage_missing)

```


```{r}

total_rows <- nrow(books)
missing_values <- colSums(is.na(books))
percentage_missing <- (missing_values / total_rows) * 100

# Display the percentage of missing values for each column
print(percentage_missing)

```

3.a) Rating Distribution

```{r}
#Converting the rating column to numeric type
# Correct way to convert 'Book-Rating' to numeric
ratings$`Book-Rating` <- as.numeric(as.character(ratings$`Book-Rating`))

# Correct ggplot call
library(ggplot2)
library(scales) # For label_number()

ggplot(ratings, aes(x=`Book-Rating`)) +
  geom_histogram(binwidth=0.5, fill="blue", color="black") +
  theme_minimal() +
  labs(title="Distribution of Book Ratings", x="Rating", y="Frequency") +
  scale_y_continuous(labels = label_number())



```
Mostly users rated 0

3.b Diplaying top-10 highest average ratting per book
```{r}
# Calculate average rating per book
avg_ratings <- ratings %>%
  group_by(ISBN) %>%
  summarise(AverageRating = mean('Book-Rating'))

# Top 10 highest rated books (filtering out books with very few ratings might be necessary)
top_rated <- avg_ratings %>%
  arrange(desc(AverageRating)) %>%
  head(10)

# Bar plot for top rated books
ggplot(top_rated, aes(x=reorder(ISBN, -AverageRating), y=AverageRating)) +
  geom_bar(stat="identity", fill="coral") +
  theme_minimal() +
  labs(title="Top 10 Highest Rated Books", x="ISBN", y="Average Rating") +
  coord_flip()

```
Let us inspect the users dataframe

 1.First let us plot the percentage of missing values of columns present in this datafram

```{r}

missing_percentage <- colSums(is.na(users)) / nrow(users) * 100

missing_df <- data.frame(Column = names(missing_percentage), MissingPercentage = missing_percentage)

# Plot
ggplot(missing_df, aes(x=Column, y=MissingPercentage)) +
  geom_bar(stat="identity", fill="tomato") +
  theme_minimal() +
  labs(title="Percentage of Missing Values by Column in Users Dataset", x="Column", y="Percentage Missing") +
  coord_flip()

```
Only the age column is missing and the missing percentage is 40.

2. If we focus on the location column, we can extract the country alone and plot the histogram of frequency.

```{r}

#We know that the country is obtained by splitting based on comma and extracting the third, i.,e last string. 
users$Country <- sapply(strsplit(as.character(users$Location), ","), function(x) trimws(x[length(x)]))

 #Get the country counts
country_counts <- table(users$Country)
#Plotting only top-10 countries

# Convert to dataframe for plotting
# Convert the table of country counts to a dataframe for plotting
country_df <- as.data.frame(country_counts, stringsAsFactors = FALSE)
names(country_df) <- c("Country", "Count")

# Sorting and selecting the top 10 countries for a cleaner plot
top_countries_df <- head(country_df[order(-country_df$Count),], 10)

# Plotting the top 10 countries
ggplot(top_countries_df, aes(x=reorder(Country, Count), y=Count)) +
  geom_bar(stat="identity", fill="steelblue") +
  geom_text(aes(label=Count), position=position_dodge(width=0.9), hjust=-0.1, vjust=0.5) +
  theme_minimal() +
  labs(title="Top 10 Countries by User Count", x="Country", y="Count") +
  coord_flip()
```
Now let us plot the age distribution:
```{r}
# Plotting the age distribution
ggplot(users, aes(x=Age)) +
  geom_histogram(binwidth=5, fill="cornflowerblue", color="black") +
  theme_minimal() +
  labs(title="Age Distribution of Users", x="Age", y="Frequency") +
  xlim(c(0, 100))

```
Let us make use of both the age and country columns to gather some visualizations

```{r}
top_countries <- users %>%
  group_by(Country) %>%
  summarise(UserCount = n()) %>%
  arrange(desc(UserCount)) %>%
  top_n(10, UserCount) %>%
  pull(Country)

# Step 2: Filter the dataset to include only users from the top 10 countries
users_top_countries <- users %>%
  filter(Country %in% top_countries)

# Step 3: Create the box plot
ggplot(users_top_countries, aes(x=Country, y=Age)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title="Age Distribution in Top 10 Countries", x="Country", y="Age") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
```


Considering the issues identified with the age column, we have decided to exclude it from further analysis for the following reasons:

1. Over 40% of the data in the age column are missing, which significantly undermines its reliability and the comprehensiveness of any analysis derived from it.

2. The presence of extreme outliers, specifically age values greater than 120, further compromises the integrity of the dataset and could lead to skewed analyses and interpretations.

Now let us inspect the dataframe concerning books

As usual let us begin by inspecting the missing values in this dataframe.

```{r}

missing_percentage <- sapply(books, function(x) sum(is.na(x)) / length(x) * 100)
missing_percentage
```

No missing values present!

Let us now plot the top-20 authors by number of books

```{r}
top_authors <- books %>%
  filter(!is.na(`Book-Author`)) %>%  # Corrected reference to "Book-Author"
  group_by(`Book-Author`) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  top_n(20, Count) %>%
  arrange(desc(Count))

# Now, plotting the corrected data
ggplot(top_authors, aes(x=reorder(`Book-Author`, Count), y=Count)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=Count), position=position_stack(vjust=0.5), color="white") +
  theme_minimal() +
  coord_flip() +
  labs(title="Top 20 Authors by Number of Books", x="Author", y="Number of Books")
```

Plotting the top 20 publishers


```{r}

top_publishers <- books %>%
  group_by(Publisher) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  top_n(20, Count) %>%
  arrange(desc(Count))

# Plotting
ggplot(top_publishers, aes(x=reorder(Publisher, Count), y=Count)) +
  geom_bar(stat="identity", fill="coral") +
  geom_text(aes(label=Count), position=position_stack(vjust=0.5), color="white") + # Adding text labels inside the bars
  theme_minimal() +
  coord_flip() +
  labs(title="Top 20 Publishers by Number of Books", x="Publisher", y="Number of Books")

```

Let us look at the year trend of publication

```{r}
# Group by Year of Publication and count the number of books
books_filtered <- books %>%
  filter(`Year-Of-Publication` >= 1950 & `Year-Of-Publication` <= 2010) %>%
  mutate(`Year-Of-Publication` = as.numeric(as.character(`Year-Of-Publication`))) # Ensure the year is numeric

# Count the number of books published each year
year_count <- books_filtered %>%
  group_by(`Year-Of-Publication`) %>%
  summarise(Count = n()) %>%
  arrange(`Year-Of-Publication`)

# Plotting the trend of book publications over the years
ggplot(year_count, aes(x=`Year-Of-Publication`, y=Count)) +
  geom_line(color="steelblue") +  # Draw the trend line
  geom_point(color="darkred", size=2, alpha=0.6) +  # Add points for each year
  theme_minimal() +  # Use a minimal theme
  labs(title="Year of Publication Count (1950 - 2010)",
       x="Year", y="Total Books Published") +
  xlim(1950, 2010)  # Set the x-axis limits

```

The trend observed in the publication data is interesting, revealing a sudden spike in book publications post-1980, with an all-time high around the year 2000. This surge can likely be attributed to the advent of the internet and the emergence of e-commerce platforms such as Amazon, which significantly boosted the book industry. Subsequently, there appears to be a notable decline in publications after roughly 2005. This decrease might be due to the introduction and increasing popularity of alternative media formats, such as video and audiobooks, which diversified consumers' options for content consumption.

EDA Part-2:

Now let us combine two datasets: ratings and books and ratings and users separately and try answering the following questions:

1. Authors trend with the highest rated.
2. Publisher trend with the highest rated.
3. Books trend with the most rated.


Now let us combine Books and Ratings dataframe together to draw some aforementioned insights

```{r}
#Merge both the dataframes

# Left join 'books' with 'ratings' on the 'ISBN' column
books_ratings <- merge(books, ratings, by = "ISBN")
```

1. Let us plot the top-20 average rating

```{r}
# Get the top 20 authors by book count
authors_top20 <- books %>%
  count(`Book-Author`, sort = TRUE) %>%
  top_n(20, n) %>%
  pull(`Book-Author`)

# Filter 'books' to include only the top 20 authors
books_top20 <- books %>%
  filter(`Book-Author` %in% authors_top20)

# Excluding NA ratings
books_with_ratings <- left_join(books_top20, ratings, by = "ISBN") %>%
  filter(!is.na(`Book-Rating`), `Book-Rating` != 0)

# Calculate the average rating for the top 20 authors
top20_author_ratings <- books_with_ratings %>%
  group_by(`Book-Author`) %>%
  summarise(AverageRating = mean(`Book-Rating`, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(AverageRating))

# Display the top 20 authors with their average rating
#top20_author_ratings
# Plot the average ratings for the top 20 authors
ggplot(top20_author_ratings, aes(x=reorder(`Book-Author`, AverageRating), y=AverageRating, fill=AverageRating)) +
  geom_bar(stat="identity") +
   geom_text(aes(label=round(AverageRating,1)), 
            position=position_dodge(width=0.9),  # Adjust position to fit within the bars
            hjust=1.1,                           # Adjust 
            color="black",                        # Text color
            size=3) + 
  coord_flip() +  # Flip the plot to make it horizontal
  theme_minimal() +
  scale_fill_gradient(low = "blue", high = "red") +  # Optional: color gradient based on rating
  labs(title="Average Rating of Top 20 Authors", x="Author", y="Average Rating")
  
```
2. Publisher trend with the highest rated.

```{r}
## Identify the top 20 publishers by book count
publishers_top20 <- books %>%
  count(Publisher, sort = TRUE) %>%
  top_n(20, n) %>%
  pull(Publisher)

top20_publishers_books <- books_ratings %>%
  filter(Publisher %in% publishers_top20 & `Book-Rating` > 0)
# Calculate average rating for each of the top 20 publishers
top20_publishers_avg_rating <- top20_publishers_books %>%
  group_by(Publisher) %>%
  summarise(AverageRating = mean(`Book-Rating`, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(AverageRating))
# Plot the average ratings for the top 20 publishers with numbers on the bars
ggplot(top20_publishers_avg_rating, aes(x=reorder(Publisher, AverageRating), y=AverageRating)) +
  geom_bar(stat="identity", fill="lightblue") +
  geom_text(aes(label=round(AverageRating, 1)), position=position_dodge(width=0.9), hjust=-0.1, vjust=0, color="black", size=3) +
  coord_flip() +
  theme_minimal() +
  labs(title="Average Rating of Books from Top 20 Publishers", x="Publisher", y="Average Rating")

```

3. Books with the highest average ratings

```{r}
# Identify the top 20 most frequently occurring books
# Get the top 20 books by occurrence in the dataset
books_top20 <- books %>%
  count(`Book-Title`, sort = TRUE) %>%
  top_n(20, n) %>%
  pull(`Book-Title`)

# Filter the merged dataset for these top 20 books and exclude zero ratings
top20_books <- books_ratings %>%
  filter(`Book-Title` %in% books_top20 & `Book-Rating` > 0)

# Calculate average ratings for these top 20 books
top20_books_avg_rating <- top20_books %>%
  group_by(`Book-Title`) %>%
  summarise(AverageRating = mean(`Book-Rating`, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(AverageRating))

# Now select the top 10 books based on their average rating
top10_books <- head(top20_books_avg_rating, 10)

# Plot the average ratings for the top 10 books
ggplot(top10_books, aes(x=reorder(`Book-Title`, AverageRating), y=AverageRating)) +
  geom_col(fill="lightblue") +
  geom_text(aes(label=sprintf("%.1f", AverageRating)), position=position_dodge(width=0.9), hjust=-0.1, vjust=0, color="black", size=3) +
  coord_flip() +
  theme_minimal() +
  labs(title="Average Rating of Top 10 Books", x="Book Title", y="Average Rating")
```


